Using best practices for forecasting biodiversity changes in breeding birds

### 2017-08-01 manuscript/draft.Rmd
more

Biodiversity forecasts are important for conservation, management, and
evaluating how well current models characterize natural systems. While there are
an increasing number of studies forecasting biodiversity, there is little
information available on how well these forecasts work. Most diversity forecasts
do not conduct meaningful evaluations of the forecasts, fail to account for
uncertainty, and do not use time-series data that captures the actual dynamics
being studied. Using hindcasting we made and evaluated `r numbers$n_forecasts`
forecasts for breeding bird species richness at `r numbers$N_predict_sites`
sites throughout the continental United States, following ten suggested best
practices for forecasting. Forecasts were made for up to a decade after the
final training observations. While all of the models we evaluated produce
reasonable forecasts, none of them consistently outperforms the simple baseline
of constant long-term average richness. At present, our best forecast for future
richness is thus to assume that it is unchanging, and current modeling
approaches do not reliably improve on this simple baseline. Most of the best
practices implemented in this study directly influence the outcome of the
forecasts and result in different conclusions from those made in previous
studies. This emphasizes the importance of following good practice in making
ecological forecasts and of starting to actively make more forecasts in such a
way that they can be quickly evaluated and improved on.

@@

In contrast to the time-series models, most attempts to predict species richness
focus on using correlative models to predict richness changes over time from
environmental data. We tested three common variants of this approach - direct
modeling of species richness, stacking individual species distribution models,
and joint species distribution models (JSDMs). Following the standard approach,
site-level random effects were not included in these models as predictors,
meaning that this approach implicitly assumes that two sites with identical
Bioclim, elevation, and NDVI values should have identical richness
distributions. As above, we included observer effects & the associated
uncertainty by running these models 500 times (once per MCMC sample).

**"Macroecological" model.** We used a boosted regression tree model using the
`gbm` package [@ridgeway2017] to directly model species richness as a function
of environmental variables. Boosted regression trees are a form of tree based
modeling that work by fitting thousands of small tree-structured models
sequentially, with each tree optimized to reduce the error of its predecessors.
They are flexible models that are considered well suited for prediction
[@elith2008]. This model was optimized using a Gaussian likelihood, with a
maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000 trees. The
number of trees used for prediction was selected using the "out of bag"
estimator; this number averaged [[N]] for the non-observer data and [[N]] for
the observer-corrected data.

@@

**Joint Species Distribution Models.** Joint species distribution models (JSDMs)
are a new approach that makes predictions about the full composition of a
community instead of modeling each species independently as above. JSDMs remove
the assumed independence among species and explicitly account for the
possibility that a site will be much more (or less) suitable for birds in
general, or particular groups of birds, than one would expect based on the
available environmental measurements alone. As a result, JSDMs do a better job
of representing our uncertainty about richness, while stacked SDMs underestimate
it [@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
[@harris2015] because it is the only JSDM that describes species' environmental
associations with nonlinear functions.

@@

Variance in species richness within sites was lower than predicted by several
common count models, such as the Poisson or binomial (i.e. richness was
underdispersed for individual sites), so these count models would have had
difficulty fitting the data. Since richness had a relatively large mean ([[N]])
and all models already produce continuous richness estimates (due to summing
probabilities in the (J)SDM case), we defined model performance for all models
in terms of continuous Gaussian errors instead of using discrete count
distributions. When a model was run multiple times for the purpose of correcting
for observer effects, we used the mean of those runs' point estimates as our
final point estimate and we calculated the uncertainty using the law of total
variance (i.e.  the average of the model runs' variance, plus the variance in
the point estimates).

We evaluated the performance of each model for forecasting using the data for
each year between `r timeframe$last_train_year` and `r timeframe$end_yr`. We used
three metrics for evaluating performance: 1) root-mean-square error (RMSE) to
determine how far, on average, the models' predictions were from the observed
value; 2) the 95% prediction interval coverage to determine how well the models
predicted the range of possible outcomes; and 3) deviance (i.e. negative 2 times
the Gaussian log-likelihood) as an integrative measure of fit incorporating good
point estimates, precision, and coverage.  In addition to looking at forecast
performance in general we evaluated how performance changed as the time horizon
of forecasting increased by plotting performance metrics against the forecast
horizon.

@@

The site-observer model found that 70% of the variance in richness in the
training set could be explained by differences among sites, and 21% could be
explained by differences among observers. The remaining 9% represents residual
variation, where a given observer might report a different number of species in
different years. In the training set, the residuals had a standard deviation of
about [[+/- 5]] species. After correcting for observer differences, there was
little temporal autocorrelation in these residuals ([[about 0.04, +/-
something]]), suggesting that richness is approximately stationary on these time
scales.

When comparing forecasts for richness across sites all methods performed well
(Figure 4; all $R^2 > 0.5$). However SDMs (both stacked and joint) and the
macroecological model all failed to successfully forecast the highest-richness
sites resulting in a notable clustering of predicted values near ~60 species and
the poorest performance ($R^2$ `r numbers$env_R2s`, versus XX-XX for the
within-site methods).

@@

While the models generally did well in absolute terms (Figure 4), none tended to
outperform the "average" baseline from the generalized linear mixed model
(Figure 4A). [[Stats go here]].  The auto-ARIMA was generally the
best-performing non-baseline model, but in many cases (67% of the time), the
best-fitting ARIMA model chosen by the model selection procedure had no
autoregressive, drift, or moving average terms, making it similar to the
"Average" model.

![](../figures/model_violins.png)

@@

Most models produced confidence intervals that were too narrow, indicating
overconfident predictions (Figure 4B). The random forest-based SDM stack was the
most overconfident model, with only `r numbers$rf_coverage_pct`% of observations
falling inside its 95% confidence intervals. This stacked SDM's narrow
predictive distribution also caused it to have notably higher deviance (Figure
4C) than the next-worst model, even though its point estimates were not
unusually bad in terms of RMSE (3A). As discussed elsewhere [@harris2015], this
overconfidence is a product of the assumption in stacked SDMs that errors in the
species-level predictions are independent. Because they avoided this assumption,
the GBM-based "macroecological" model and especially the mistnet JSDM had
reasonably well-calibrated uncertainty estimates (Figure 4B); as a result, their
relative performance was higher in terms of deviance than in terms of RMSE. The
"naive" model was the only model whose confidence intervals were too wide
(Figure 4B), which can be attributed to the rapid rate at which these intervals
expand as the time lag increases (Figure 1).

@@

![Accounting for differences between observers typically has a small impact on
accuracy (especially for the "naive" model), but occasonally changes the
predicted richness values by 10+ species. These changes usually improve the
model's predictions, but not always.](../figures/observers.png)

@@

Forecasting is an emerging imperative in ecology and as such the field needs to
develop and follow best practices for conducting and evaluating ecological
forecasts. We have laid out ideas for a number of these practices (Box 1) and
attempted to implement them in a single study that builds and evaluates
forecasts of biodiversity in the form of species richness. The results of this
effort are both promising and humbling. On an absolute basis, when comparing
forecasts across sites, many different approaches to forecasting produce
reasonable forecasts. If a site is predicted to have a high number of species in
the future, relative to other sites, it generally does. However, the forecasts
perform relatively poorly at predicting variation in richness over time at a
location. If a site is predicted to have a higher richness next year than it did
last year there is little confidence that this will be the case. As a result,
the baseline models, like the long-term average, along with site-level
time-series models, provided some of the best forecasts for future biodiversity.

The most commonly used method to forecast future biodiversity, stacked SDMS
[@urban2015], provided the worst overall forecasts of all methods evaluated. In
general, SDM-based forecasts are rarely evaluated, particularly in terms of
repeated observations at the same site. Since site-level richness is relatively
stable, model evaluations that show high accuracy in predicting differences
among sites may not indicate much about a model's ability to predict changes
over time. This issue has also recently been highlighted as an issue
for using SDMs to predict the future locations of individual species
[@rapacciuolo2012]. This result is particularly sobering because this approach
forms the main foundation for estimates of the predicted loss of biodiversity to
climate change [@thomas2004; @urban2015]. The poor performance of SDM-based
forecasts highlights the crucial importance of comparing multiple modeling
approaches when conducting ecological forecasts, and in particular the value of
comparing results to simple baselines to avoid over-interpreting the information
present in these forecasts (Box 1).  Disciplines that have more mature forecasting
cultures often do this by reporting "forecast skill", which is the improvement
in the forecast relative to a simple baseline [@jolliffe2003]. We recommend
following Ye et al. [@ye2015] and adopting this approach in future ecological
forecasting research. It also highlights the need for using hindcasting and
time-series data for assessing the effectiveness of forecasting methods, since
this allows comparisons to baselines and assessment of predicted dynamics within
sites or regions.

One of the main alternatives to summed SDMs is modeling richness directly using
environmental variables and using the resulting models to make forecasts
[@algar2009; @distler2015]. Two comparisons of the the SDM and direct richness
modeling approaches reported that the methods yielded equivalent results for
forecasting diversity. While our results generally support rough equivalence
between stacked SDMs and direct richness modeling for point estimates, they also
show that modeling richness directly results in much better estimation of the
uncertainty of the forecast and therefore is a better overall approach to
forecasting richness. A similar result is seen when comparing joint species
distribution models (JSDMs) to stacked single species distribution models. For
point estimates the joint distribution models are roughly equivalent to stacked
SDMs, but the JSDMs provide much better estimates of uncertainty [@harris2015;
others?]. In fact, JSDMs and modeling richness directly provide some of the best
estimates of uncertainty across all modeling approaches. This highlights the
importance of evaluating models based on their uncertainty as well as their mean
estimates (Box 1). Not doing so is particularly problematic in this case because the
summed species distribution models are overly confident meaning that using them
would lead to believing that richness would be restricted to a much narrower
range than would actually occur.

@@

It is also possible that models that include species' relationships to their
environments or direct environmental constraints on richness may continue to
produce forecasts that are worse than simply assuming the systems are
static. This would be expected to occur if richness in these systems is not
changing over the relevant time scales and therefore simple models of no
directional change are most appropriate. Recent suggestions that local scale
richness in some systems is not changing directionally at multi-decadal scales
supports this possibility [@brown2001; @ernest2001; @dornelas2014;
@vellend2013]. This lack of change may be expected even in the presence of
substantial changes in environmental conditions and species composition at a
site due to the replacement of species from the regional pool [@brown2001;
@ernest2001]. On average the Breeding Bird Survey sites used in this study show
little change in richness ([[Add summary stat here]]) [see also
@lasorte2005]. This absence of rapid change is beneficial for the absolute
accuracy of forecasts across sites, because if a past year's richness is known
it is easy to estimate a future richness, and explains why stacked SDMs perform
relatively well at this task despite failing to capture meaningful
dynamics. However it makes it difficult to improve forecasts relative to simple
baselines, since those baselines are close to representing what is actually
occurring in the system. This suggests that simple time-series models and
baselines should be actively considered for forecasts of richness and other
stable aspects of biodiversity and also suggests that future efforts to
understand and forecast biodiversity should also focus on composition since it
is expected to be more dynamic [@ernest2001; @dornelas2014].

>>>>

Biodiversity forecasts are important for conservation, management, and
evaluating how well current models characterize natural systems. While there are
an increasing number of studies forecasting biodiversity, there is little
information available on how well these forecasts work. Most diversity forecasts
do not conduct meaningful evaluations of the forecasts, fail to account for
uncertainty, and do not use time-series data that captures the actual dynamics
being studied. Using hindcasting we made and evaluated `r numbers$n_forecasts`
forecasts for breeding bird species richness from each of 6 modeling approaches
at `r numbers$N_predict_sites` sites throughout the continental United States,
following ten suggested best practices for forecasting. Forecasts were made for
up to a decade after the final training observations. While all of the models we
evaluated produce reasonable forecasts, none of them consistently outperforms
the simple baseline of constant long-term average richness. At present, our best
forecast for future richness is thus to assume that it is unchanging, and
current modeling approaches do not reliably improve on this simple baseline.
Most of the best practices implemented in this study directly influence the
outcome of the forecasts and result in different conclusions from those made in
previous studies. This emphasizes the importance of following good practice in
making ecological forecasts and of starting to actively make more forecasts in
such a way that they can be quickly evaluated and improved on.

@@

In contrast to the site-level time-series models, most attempts to predict
species richness focus on using correlative models based on environmental
variables. We tested three common variants of this approach: direct modeling of
species richness; stacking individual species distribution models; and joint
species distribution models (JSDMs). Following the standard approach, site-level
random effects were not included in these models as predictors, meaning that
this approach implicitly assumes that two sites with identical Bioclim,
elevation, and NDVI values should have identical richness distributions. As
above, we included observer effects & the associated uncertainty by running
these models 500 times (once per MCMC sample).

**"Macroecological" model.** We used a boosted regression tree model using the
`gbm` package [@ridgeway2017] to directly model species richness as a function
of environmental variables. Boosted regression trees are a form of tree based
modeling that work by fitting thousands of small tree-structured models
sequentially, with each tree optimized to reduce the error of its predecessors.
They are flexible models that are considered well suited for prediction
[@elith2008]. This model was optimized using a Gaussian likelihood, with a
maximum interaction depth of 5, shrinkage of 0.015, and up to 10,000 trees. The
number of trees used for prediction was selected using the "out of bag"
estimator; this number averaged `r numbers$n_gbm[1]` for the non-observer data
and `r numbers$n_gbm[2]` for the observer-corrected data.

@@

**Joint Species Distribution Models.** Joint species distribution models (JSDMs)
are a new approach that makes predictions about the full composition of a
community instead of modeling each species independently as above. JSDMs remove
the assumed independence among species and explicitly account for the
possibility that a site will be much more (or less) suitable for birds in
general (or particular groups of birds) than one would expect based on the
available environmental measurements alone. As a result, JSDMs do a better job
of representing our uncertainty about richness, while stacked SDMs underestimate
it [@harris2015; [[other refs, e.g., Clark]]]. We used the `mistnet` package
[@harris2015] because it is the only JSDM that describes species' environmental
associations with nonlinear functions.

@@

Variance in species richness within sites was lower than predicted by several
common count models, such as the Poisson or binomial (i.e. richness was
underdispersed for individual sites), so these count models would have had
difficulty fitting the data. Since richness had a relatively large mean 
(`r numbers$richness_summary$mean`) and all models already produce continuous
richness estimates (due to summing probabilities in the (J)SDM case), we defined
model performance for all models in terms of continuous Gaussian errors instead
of using discrete count distributions. When a model was run multiple times for
the purpose of correcting for observer effects, we used the mean of those runs'
point estimates as our final point estimate and we calculated the uncertainty
using the law of total variance (i.e.  the average of the model runs' variance,
plus the variance in the point estimates).

We evaluated each model's for forecasts using the data for each year between `r
timeframe$last_train_year` and `r timeframe$end_yr`. We used three metrics for
evaluating performance: 1) root-mean-square error (RMSE) to determine how far,
on average, the models' predictions were from the observed value; 2) the 95%
prediction interval coverage to determine how well the models predicted the
range of possible outcomes; and 3) deviance (i.e. negative 2 times the Gaussian
log-likelihood) as an integrative measure of fit incorporating good point
estimates, precision, and coverage.  In addition to looking at forecast
performance in general we evaluated how performance changed as the time horizon
of forecasting increased by plotting performance metrics against the forecast
horizon.

@@

The site-observer model found that 70% of the variance in richness in the
training set could be explained by differences among sites, and 21% could be
explained by differences among observers. The remaining 9% represents residual
variation, where a given observer might report a different number of species in
different years. In the training set, the residuals had a standard deviation of
about `r numbers$resid_sd` species. After correcting for observer differences,
there was little temporal autocorrelation in these residuals ([[about 0.04, +/-
something]]), suggesting that richness is approximately stationary on these time
scales.

When comparing forecasts for richness across sites all methods performed well
(Figure 4; all $R^2 > 0.5$). However SDMs (both stacked and joint) and the
macroecological model all failed to successfully forecast the highest-richness
sites resulting in a notable clustering of predicted values near ~60 species and
the poorest performance ($R^2$=`r numbers$env_R2s`, versus $R^2$=`r numbers$ts_R2s` 
for the within-site methods).

@@

While the models generally did well in absolute terms (Figure 4), none
consistently outperformed the "average" baseline from the generalized linear
mixed model (Figure 4A). [[Stats go here]].  The auto-ARIMA was generally the
best-performing non-baseline model, but in many cases (67% of the time), the
auto.arima procedure automatically selected a model with no autoregressive,
drift, or moving average terms, making it similar to the "Average" model.

![None of the models provided a consistent improvement over the "average"
baseline, indicated by a horizontal line. **A.** Models' absolute error was
generally similar or larger than "average" model, with large outliers in both
directions. **B.** The other models' deviance was also generally higher than
the "average" baseline.](../figures/model_violins.png)

@@

Most models produced confidence intervals that were too narrow, indicating
overconfident predictions (Figure 4B). The random forest-based SDM stack was the
most overconfident model, with only `r numbers$rf_coverage_pct`% of observations
falling inside its 95% confidence intervals. This stacked SDM's narrow
predictive distribution also caused it to have notably higher deviance (Figure
4C) than the next-worst model, even though its point estimates were not
unusually bad in terms of RMSE (3A). As discussed elsewhere [@harris2015], this
overconfidence is a product of the assumption in stacked SDMs that errors in the
species-level predictions are independent. Because they avoided this assumption,
the GBM-based "macroecological" model and the mistnet JSDM had reasonably
well-calibrated uncertainty estimates (Figure 4B); as a result, their relative
performance was higher in terms of deviance than in terms of RMSE. The "naive"
model was the only model whose confidence intervals were too wide (Figure 4B),
which can be attributed to the rapid rate at which these intervals expand
(Figure 1).

@@

![Accounting for differences between observers typically has a small impact on
accuracy (especially for the "naive" model), but occasonally changes the
predicted richness values by more than 10 species. These changes usually improve
the model's predictions, but not always.](../figures/observers.png)

@@

Forecasting is an emerging imperative in ecology; as such, the field needs to
develop and follow best practices for conducting and evaluating ecological
forecasts. We have laid out ideas for a number of these practices (Box 1) and
attempted to implement them in a single study that builds and evaluates
forecasts of biodiversity in the form of species richness. The results of this
effort are both promising and humbling. On an absolute basis, when comparing
forecasts across sites, many different approaches to forecasting produce
reasonable forecasts. If a site is predicted to have a high number of species in
the future, relative to other sites, it generally does. However, the forecasts
perform relatively poorly at predicting variation in richness over time. If a
site is predicted to have a higher richness next year than it did last year,
there is little confidence that this will be the case. As a result, baseline
models, like the long-term average, along with the other site-level time-series
models, consistently provided the best forecasts for future biodiversity.

The most commonly used method to forecast future biodiversity, stacked SDMs
[@urban2015], provided the worst overall forecasts of all methods evaluated. In
general, SDM-based forecasts are rarely evaluated, particularly in terms of
repeated observations at the same site. Since site-level richness is relatively
stable, model evaluations that show high accuracy across spatial gradients may
not indicate much about a model's ability to predict changes over time. This
issue has also recently been highlighted as an issue for using SDMs to predict
the future locations of individual species [@rapacciuolo2012]. This result is
particularly sobering because this approach forms the main foundation for
estimates of the predicted loss of biodiversity to climate change [@thomas2004;
@urban2015]. The poor performance of SDM-based forecasts highlights the crucial
importance of comparing multiple modeling approaches when conducting ecological
forecasts, and in particular the value of comparing results to simple baselines
to avoid over-interpreting the information present in these forecasts (Box 1).
Disciplines that have more mature forecasting cultures often do this by
reporting "forecast skill", i.e., the improvement in the forecast relative to a
simple baseline [@jolliffe2003]. We recommend following Ye et al. [@ye2015] and
adopting this approach in future ecological forecasting research. It also
highlights the need for using hindcasting and time-series data for assessing the
effectiveness of forecasting methods, since this allows comparisons to baselines
and assessment of predicted dynamics within sites or regions.

One of the main alternatives to stacked SDMs is modeling richness directly using
environmental variables and using the resulting models to make forecasts
[@algar2009; @distler2015]. Two comparisons of the the SDM and direct richness
modeling approaches reported that the methods yielded equivalent results for
forecasting diversity. While our results generally support rough equivalence
between stacked SDMs and direct richness modeling for point estimates, they also
show that stacked SDMs dramatically understimate the uncertainty. For this
reason, direct modeling of richness therefore is a better overall approach to
forecasting richness. A similar result is seen when comparing joint species
distribution models (JSDMs) to stacked single species distribution models. For
point estimates the joint distribution models are roughly equivalent to stacked
SDMs, but the JSDMs provide much better estimates of uncertainty [@harris2015;
others?]. In fact, JSDMs and direct richness modeling provide some of the best
estimates of uncertainty across all modeling approaches. This highlights the
importance of evaluating models based on their uncertainty as well as their mean
estimates (Box 1). Not doing so is particularly problematic in this case because
the summed species distribution models are overly confident meaning that using
them would lead to believing that richness would be restricted to a much
narrower range than would actually occur.

@@

It is also possible that models that include species' relationships to their
environments or direct environmental constraints on richness may continue to
produce forecasts that are worse than simply assuming the systems are static.
This would be expected to occur if richness in these systems is not changing
over the relevant time scales and therefore simple models of no directional
change are most appropriate. Recent suggestions that local scale richness in
some systems is not changing directionally at multi-decadal scales supports this
possibility [@brown2001; @ernest2001; @dornelas2014; @vellend2013]. This lack of
change may be expected even in the presence of substantial changes in
environmental conditions and species composition at a site due to the
replacement of species from the regional pool [@brown2001; @ernest2001]. On
average the Breeding Bird Survey sites used in this study show little change in
richness (site-level SD of `r numbers$resid_sd` species, after controlling for
differences among observers) [see also @lasorte2005]. This absence of rapid
change is beneficial for the absolute accuracy of forecasts across sites,
because if a past year's richness is known it is easy to estimate a future
richness, and explains why stacked SDMs perform relatively well at this task
despite failing to capture meaningful dynamics. However it makes it difficult to
improve forecasts relative to simple baselines, since those baselines are close
to representing what is actually occurring in the system. This suggests that
simple time-series models and baselines should be actively considered for
forecasts of richness and other stable aspects of biodiversity and also suggests
that future efforts to understand and forecast biodiversity should also focus on
composition since it is expected to be more dynamic [@ernest2001;
@dornelas2014].
